A Coq Formalization of Stable Sorts, Radix Sort, and the Burrows-Wheeler Transform

### 2019-10-17 draft_paper/draft_paper.tex
Finish content

\author{Andrew Appel}
\affiliation{
  \department{Department of Computer Science}
  \institution{Princeton University}
  \city{Princeton}
  \state{New Jersey}
  \postcode{08540}
  \country{USA}
}
\email{appel@princeton.edu}

@@

There has not been much work on formally verifying stable sorts, or
rather, the stability of stable sorts. While Bird et~al.
\cite{birdmu,pearls} provide an elegant derivation and proof for the
BWT, formalizing it in Coq reveals that they rely on a few lemmas that
are not trivial to prove, including the correctness of
least-significant digit (LSD) radix sort. We develop the theory of
stable sorts, and use this theory to prove the correctness of LSD
radix sort and thereby the BWT.
\end{abstract}

@@

Stable sorts are an important and fundamental type of sorting
algorithm. A sort is stable if it preserves the relative order of
equal elements. For example, if we are stably sorting the list
$[\var{banana},\ \allowbreak\var{armpit},\ \allowbreak\var{carrot},\ \allowbreak\var{apple}]$,
we have to produce
$[\var{armpit},\ \allowbreak\var{apple},\ \allowbreak\var{banana},\ \allowbreak\var{carrot}]$
and not
$[\var{apple},\ \allowbreak\var{armpit},\ \allowbreak\var{banana},\ \allowbreak\var{carrot}]$.
Even though both are sorted on the first character, because
\var{armpit} preceded \var{apple} in the original lists, it must also
precede \var{apple} in the stably sorted list.

Stability can be directly useful in certain contexts, such as when
sorting a table by a single column. However, stability is also crucial
to the correctness of certain general-purpose algorithms. One such
algorithm, or set of algorithms, is radix sort. Radix sorts are used
to sort data lexicographically, and can be more efficient than
comparison-based sorting methods, both in theory and in practice
\cite{McIlroy93,Bentley:1997:FAS:314161.314321,10.1007/978-3-540-89097-3_3}.
Many radix sorts rely on a some stable sorting subroutine.

@@

In addition, to our knowledge no one has produced a formally verified
implementation of the BWT, by which we mean an implementation with a
proof that the inverse-transform inverts the forwards-transform. Bird
et~al.\cite{alg-of-prog,birdmu} provide an excellent derivation of the
BWT and its inverse based on the technique program calculation, in
which programs are calculated from their specifications using
equational reasoning. However, their derivation leaves a few crucial
lemmas unproven, including the correctness of LSD radix sort.

In this paper, we develop more fully the theory of stable sorts: we
give multiple definitions for permutations and stable permutations and
prove them equivalent. This allows us to choose whichever definition
makes the proof easiest when we are reasoning about stability. We also
proved that any two sorts are interchangeable.
\begin{restatable*}[StableSort\_unique]{corollary}{stablesortunique}
  \label{thm:stablesort_unique}
  For all stable sorting functions $f, g: \var{list}~A \to
  \var{list}~A$,
  \begin{equation*}
    (\forall l,f~l = g~l)
  \end{equation*}
\end{restatable*}

Using \cref{thm:stablesort_unique}, we prove the correctness of an LSD
radix sort that relies on insertion sort to sort each column. This
choice was crucial to the proof: parametrizing radix sort over any
stable sort made it difficult to reason about, but reasoning about
insertion sort is very elegant. \cref{thm:stablesort_unique} allows us
to recover the generality of parametrizing over all stable sorts,
because it proves that insertion sort is equivalent to any other
stable sort.

@@

The implementation is useful in its own right as a functional
specification for a potential verified implementation of a program
that relies on Burrows--Wheeler, such as bzip2. But more generally,
this provides evidence that Bird's approach of program calculation, in
which programs are derived from their specifications via algebraic
manipulations, is useful for proving the correctness of algorithms in
a machine-checked setting. Bird has applied this approach to greedy
algorithms, the Boyer-Moore algorithm, the Knuth-Morris-Pratt
algorithm, Sudoku solvers, arithmetic coding, the Schorr-Waite
algorithm, the Johnson-Trotter algorithm, and many more problems
\cite{pearls}.

Every definition and theorem in this paper corresponds directly to a
definition or theorem in our Coq sources, which we provide as
supplementary materials. Definitions are copied verbatim, modulo
whitespace. Theorems are rephrased for clarity, with the name of the
corresponding Coq theorem given in parentheses. The only exceptions
are \cref{sec:opt} and places where we give hypothetical definitions
to drive the explanation.

@@

We can define \var{Sorted} in two main ways: an index-based definition
(\var{SortedIx}), and an inductive definition (\var{Sorted}).
\begin{definition}{SortedIx}
  A list $l$ is sorted if for all indices $i$ and $j$,
  \begin{equation*}
    i \le j \implies l_i \le l_j
  \end{equation*}
\end{definition}

\begin{definition}{Sorted}
  \begin{align*}
    \infer[\var{Sorted\_nil}]{\var{Sorted}~[]}{}
    &&
    \infer[\var{Sorted\_cons}]{\var{Sorted}~(a :: l)}{%
      (\forall~x \in l,~a < x) & \var{Sorted}~l
    }
  \end{align*}
\end{definition}

\begin{restatable}[SortedIx\_iff]{theorem}{sortedixiff}
  For all lists $l$,
  \begin{equation*}
    \var{Sorted}~l \iff \var{SortedIx}~l
  \end{equation*}
\end{restatable}
\begin{proof}
  See \cref{appendix:proofs}.
\end{proof}

However, a sorting function cannot produce just any sorted list as its
output; the sorted list must also contain all the same elements as the
original list. To capture this, we could use the \var{Permutation}
relation defined in the Coq standard library. This relation holds on
two lists when they contain all the same elements, and is defined
inductively:
\begin{definition}[Permutation]
  \label{def:permutation}
  \begin{gather*}
    \infer[\var{perm\_nil}]{\var{Permutation}~[]~[]}{}
    \\
    \allowbreak
    \\
    \infer[\var{perm\_skip}]{\var{Permutation}~(x :: l)~(x :: l')}{%
      \var{Permutation}~l~l'
    }
    \\
    \allowbreak
    \\
    \infer[\var{perm\_swap}]{\var{Permutation}~(x :: y :: l)~(y :: x :: l)}{}
    \\
    \allowbreak
    \\
    \infer[\var{perm\_trans}]{\var{Permutation}~l~l''}{
      \var{Permutation}~l~l' & \var{Permutation}~l'~l''
    }
  \end{gather*}
\end{definition}

Before we go further, we should clarify what we mean by the word
\textit{permutation}, as it has two related meanings that are often
used interchangeably. We sometimes say that two lists are permutations
of one other if they contain all of the same elements, perhaps in a
different order, like the lists $l =
[\var{zebra},\ \allowbreak\var{appliance},\ \allowbreak\var{kite}]$
and $l' =
[\var{appliance},\ \allowbreak\var{kite},\ \allowbreak\var{zebra}]$
However, in mathematics we sometimes say that a permutation is
actually the bijection on the indices that defines the correspondence
between the two lists. In this case, we could map indices in $l$ to
indices in $l'$ by mapping $0 \mapsto 2$, mapping $1 \mapsto 0$, and mapping $2 \mapsto
0$. Formally, we say that $p$ is the permutation that relates $l$ to
$l'$ if for all indices $i$ and $j$,
\begin{equation*}
  l_i = l'_{p(i)}
\end{equation*}
For consistency with Coq's built-in \var{Permuation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
one other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$.

This suggests an alternative definition for \var{Permutation}:
\begin{definition}{PermutationEx}
  \label{def:permutation_ex}
  Two lists $l$ and $l'$ are permutations of one another when, letting
  $n = \var{length}~l$, there exists a permutation function $p$ on $n$
  elements such that $p(l) = l'$.
\end{definition}
Here we are overloading the notation $p(l)$ to denote the result of
shuffling a list~$l$ according to a permutation function~$p$. To
define this precisely, we have to first define how we will represent
permutation functions in Coq.

A permutation function on $n$ elements maps indices to indices, and so
has type $\nat_{<n} \to \nat_{<n}$. In Coq, we can represent
a finite function as a list of its outputs: $f : \nat_{<n} \to A$
is represented by the list $[f(0),~f(1),\allowbreak\ldots,~f(n-1)]$.
Applied to permutations, this means that the permutation~$p$ mentioned
above would be represented by $[2,~1,~0]$. This is consistent with
``one-line'' notation for permutations, in which a permutation on $n$
elements is represented by its result when applied to the
list~$[0,~2,\ldots,n-1]$. Now we can make precise the notation
$p(l)$\footnote{In Coq, this corresponds the the \var{apply} function,
  whose definition is slightly more complex due to the extra default
  parameter for \var{nth}.}
\begin{equation*}
  p(l) \coloneq \var{map}~(\lambda i. l_i)~p
\end{equation*}

However, not any list of natural numbers represents a valid
permutation function, because a permutation must be a bijection. We
can define the property $\var{PermFun}~n~p$, which holds when $p :
\var{list}~\nat$ represents a permutation function on lists of
length $n$:
\begin{definition}[PermFun]
  A list~$p$ of natural numbers is a permutation function on $n$
  elements if for all natural numbers $i$, $i \in p \iff i < n$, and if
  $p$ contains no duplicate elements\footnote{For this we can use the
    \var{NoDup} predicate from Coq's standard library.}.
\end{definition}

@@

With this we can define the specification for sorting functions:
\begin{definition}[Sort]
  A function $f : \var{list}~A \to \var{list}~A$ is a sorting function
  when for all inputs $l$,
  \begin{equation*}
    \var{Sorted}~(f~l) \land \var{Permutation}~(f~l)~l
  \end{equation*}
\end{definition}

@@

Now that we've defined sorts, we can start to consider stable sorts.
Stable sorts must preserve the relative orders of ``equivalent''
elements. This equivalence comes from the ordering we are using: any
total decidable preorder also give us a decideable equivalence
relation $\equiv$, where $x \equiv y$ when $x \le y$ and $x \ge y$.

@@

To define stable permutations, we can tweak each of the definitions of
permutation mentioned above. For example, we can modify our
function-based definition.

@@

We can also modify the inductive definition of permutations to respect
equivalence. The problematic rule is \var{perm\_swap} because we could
porentially swap two equivalent elements. We will restrict it, so that
you can only swap $x$ and $y$ when $x \nequiv y$.

@@

Finally, we can say that if two lists are in a stable permutation, and
we examine only the the elements equivalent to $x$ for some $x$, then
those elements should appear in the same order in both lists.
\begin{definition}[StablePerm]
\item Two lists $l$ and $l'$ are stable permutations of one another if
  for every $x$,
  \begin{equation*}
    \var{filter}~(\lambda y. y \equiv x)~l = \var{filter}~(\lambda y. y \equiv x)~l'
  \end{equation*}
\end{definition}
In a sense, this is a stricter version of
\cref{def:permutation_count}, because that can be rephrased as
$\var{filter}~(\lambda y. y = x)~l = \var{filter}~(\lambda y. y = x)~l'$.

@@

This means that any two stable sorts must produce the same results for
every input, and are thus interchangeable.
\stablesortunique*

This can be generalized a little further: even if a sorting algorithm
itself is not a stable sorting algorithm, if the ordering we are using
doesn't allow us to distinguish between equivalent elements, then we
will not be able to tell if equivalent elements are being swapped.
More precisely, notice that it isn't always true that $x \equiv y$ implies
$x = y$. For example, if we are sorting strings by their first
character, then $\text{\texttt{apple}} \equiv \text{\texttt{armpit}}$ under
this ordering, but clearly $\text{\texttt{apple}} \neq
\text{\texttt{armpit}}$. When we do have that the equivalence ($\equiv$)
implies Leibniz equality ($=$), we say that we have an \textit{order},
as we have when we are sorting strings using all of their characters.
Under an order, as opposed to a preorder, we can prove that every
permutation is a stable permutation and every sort is a stable sort,
which implies that every sort is equivalent to every other sort.

@@

LSD radix sort is a simple radix sort that is used to
lexicographically sort strings of equal length. We can define what it
means to sort lexicographically:

\begin{definition}[lex\_le]
  Given a preorder~$\le$ on a type $A$, we can define another preorder,
  $\lexle$ on $\var{list}~A$.
  \begin{gather*}
    \infer[\var{lex\_le\_nil}]{[] \lexle ys}{}
    \\
    \\
    \infer[\var{lex\_le\_cons\_lt}]{(x :: xs) \lexle (y :: ys)}{%
      x < y
    }
    \\
    \\
    \infer[\var{lex\_le\_cons\_eq}]{(x :: xs) \lexle (y :: ys)}{%
      x \equiv y & xs \lexle ys
    }
  \end{gather*}
\end{definition}

LSD radix sort operates by conceptually placing all the strings in a
matrix, and by stably sorting on each column, from the end of the
string to the front (\cref{fig:lsd_radixsort}). This works because if
two strings match on a column, the stable sort ``defers'' to the
preexisting ordering, and that preexisting ordering was determined by
sorting by the tail of the strings. This mirrors the way a
lexicographic ordering ``defers'' to the rest of a string when two
strings match on earlier characters.

\begin{figure}
  \centering
  \begin{tt}
    \setlength{\tabcolsep}{0pt}
    \begin{tabular}{c}
    fit \\
    rib \\
    fly \\
    vat \\
    age \\
    mix \\
    row \\
    fog
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    ri&b& \\
    ag&e& \\
    fo&g& \\
    fi&t& \\
    va&t& \\
    mi&x& \\
    fl&y& \\
    ro&w&
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    v&a&t \\
    a&g&e \\
    r&i&b \\
    f&i&t \\
    m&i&x \\
    f&l&y \\
    f&o&g \\
    r&o&w
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &a&ge \\
    &f&it \\
    &f&ly \\
    &f&og \\
    &m&ix \\
    &r&ib \\
    &r&ow \\
    &v&at
    \end{tabular}
  \end{tt}
  \caption{LSD radix in-place}
  \label{fig:lsd_radixsort}
\end{figure}

In a functional setting, we can define an LSD radixsort elegantly by
repeatedly moving the last column to the front of the matrix and
sorting on the first column (\cref{fig:lsd_radixsort_func}). We will
take as a parameter the number of columns $n$.
\begin{equation*}
  \var{radixsort}~m~n = (\var{hdsort} \comp \var{map}~\var{rrot})^n~m
\end{equation*}

\begin{figure}
  \centering
  \begin{tt}
    \setlength{\tabcolsep}{0pt}
    \begin{tabular}{c}
    fit \\
    rib \\
    fly \\
    vat \\
    age \\
    mix \\
    row \\
    fog
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c}
    tfi \\
    bri \\
    yfl \\
    tva \\
    eag \\
    xmi \\
    wro \\
    gfo
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &b&ri \\
    &e&ag \\
    &g&fo \\
    &t&fi \\
    &t&va \\
    &x&mi \\
    &y&fl \\
    &w&ro
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    ibr \\
    gea \\
    ogf \\
    itf \\
    atv \\
    ixm \\
    lyf \\
    owr
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &a&tv \\
    &g&ea \\
    &i&br \\
    &i&tf \\
    &i&xm \\
    &l&yf \\
    &o&gf \\
    &o&wr
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    vat \\
    age \\
    rib \\
    fit \\
    mix \\
    fly \\
    fog \\
    row
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &a&ge \\
    &f&it \\
    &f&ly \\
    &f&og \\
    &m&ix \\
    &r&ib \\
    &r&ow \\
    &v&at
    \end{tabular}
  \end{tt}
  \caption{LSD radix using rotations}
  \label{fig:lsd_radixsort_func}
\end{figure}

Here \var{hdsort} is an insertion sort that only examines the first
character of each string. We can implement this by defining orderings
that compare elements using some key function:

@@

For both properties, our approach will be to prove some invariant that
holds after $j$ iterations. For \var{StablePerm}, we can say that
after $j$ iterations of rotating and sorting, the matrix should be a
stable permutation of the original matrix had we rotated $j$ times.

>>>>

\author{Andrew W. Appel}
\affiliation{
  \department{Department of Computer Science}
  \institution{Princeton University}
  \city{Princeton}
  \state{New Jersey}
  \postcode{08540}
  \country{USA}
}
\email{appel@princeton.edu}

@@

There has not been much work on formally verifying stable sorts, or
rather, the stability of stable sorts. While Bird
et~al.\ \cite{birdmu,pearls} provide an elegant derivation and proof
for the BWT, formalizing it in Coq reveals that they rely on a few
lemmas that are not trivial to prove, including the correctness of
least-significant digit (LSD) radix sort. We develop the theory of
stable sorts, and use this theory to prove the correctness of LSD
radix sort and thereby the BWT.
\end{abstract}

@@

Stable sorts are an important and fundamental type of sorting
algorithm. A sort is stable if it preserves the relative order of
equal elements. For example, if we are stably sorting the list
$[\var{mix},\ \allowbreak\var{fog},\ \allowbreak\var{age},\ \allowbreak\var{fit}]$
by the first character of each word, we have to produce
$[\var{age},\ \allowbreak\var{fog},\ \allowbreak\var{fit},\ \allowbreak\var{mix}]$
and not
$[\var{age},\ \allowbreak\var{fit},\ \allowbreak\var{fog},\ \allowbreak\var{mix}]$.
Even though both are sorted on the first character, because \var{fog}
preceded \var{fit} in the original list, it must also precede
\var{fit} in the stably sorted list.

Stability can be directly useful in certain contexts, such as when
sorting a table by a single column. However, stability is also crucial
to the correctness of certain general-purpose algorithms. One such
algorithm, or set of algorithms, is radix sort. Radix sorts are used
to sort data lexicographically, and can be more efficient than
comparison-based sorting methods, both in theory and in practice
\cite{McIlroy93,Bentley:1997:FAS:314161.314321,10.1007/978-3-540-89097-3_3}.
Most radix sorts rely on a some stable sorting subroutine.

@@

In addition, to our knowledge no one has produced a formally verified
implementation of the BWT. Bird et~al.\ \cite{alg-of-prog,birdmu}
provide an excellent derivation of the BWT and its inverse based on
the technique of program calculation, in which programs are calculated
from their specifications using algebraic transformations. However,
their derivation leaves a few crucial lemmas unproven, including the
correctness of LSD radix sort.

In this paper, we develop more fully the theory of stable sorts: we
give multiple definitions for permutations and stable permutations and
prove them equivalent. This allows us to choose whichever definition
makes the proof easiest. We also prove that any two sorts are
interchangeable.
\begin{restatable*}[StableSort\_unique]{corollary}{stablesortunique}
  \label{thm:stablesort_unique}
  For all stable sorting functions $f, g: \var{list}~A \to
  \var{list}~A$,
  \begin{equation*}
    (\forall l,f~l = g~l)
  \end{equation*}
\end{restatable*}

Throughout our proofs about LSD radix sort and the BWT, we reason
about sorting algorithms whose implementation is irrelevent as long as
they are correct. We could, in principle, rely only on the fact that
they fulfill the specification of a sorting algorithm
(\cref{def:sort}). However, in practice this is difficult, because the
inductive definition of permutations can produce unhelpful induction
hypotheses, especially in the transitive case. Insetad, we reason
about specifically insertion sort. Therefore
\cref{thm:stablesort_unique} is quite useful, because we can reason
using insertion sort, and later on replace it with any other
(presumably faster) stable sort.

@@

The implementation is useful in its own right as a functional
specification for a potential verified implementation of a program
that relies on Burrows--Wheeler, such as bzip2. But more generally,
this provides evidence that Bird's approach of program calculation is
useful for proving the correctness of algorithms in a machine-checked
setting. Bird has applied this approach to greedy algorithms, the
Boyer-Moore algorithm, the Knuth-Morris-Pratt algorithm, Sudoku
solvers, arithmetic coding, the Schorr-Waite algorithm, the
Johnson-Trotter algorithm, and many more problems \cite{pearls}.

Every definition and theorem in this paper corresponds directly to a
definition or theorem in our Coq sources, which we provide as
supplementary materials. Definitions are copied verbatim, modulo
whitespace. Theorems and definitions are rephrased for clarity, with
the name of the corresponding Coq theorem given in parentheses. The
only exceptions are \cref{sec:opt} and places where we give
hypothetical definitions to drive the explanation.

@@

We can define \var{Sorted} in two main ways: an index-based definition
(\var{SortedIx}), and an inductive definition (\var{Sorted}).

\begin{definition}[SortedIx]
  A list $l$ is sorted if for all indices $i$ and $j$,
  \begin{equation*}
    i \le j \implies l_i \le l_j
  \end{equation*}
\end{definition}

We use the notation $l_i$ for the $i$th element of a list, though in
Coq, that is given by $\var{nth}~i~l~d$, where $d:A$ is a default
value in case $i \ge \var{length}~l$.

\begin{definition}[Sorted]
  \label{def:sorted}
  \begin{gather*}
    \infer[\var{Sorted\_nil}]{\var{Sorted}~[]}{}
    \\
    \infer[\var{Sorted\_cons}]{\var{Sorted}~(a :: l)}{%
      (\forall~x \in l,~a < x) & \var{Sorted}~l
    }
  \end{gather*}
\end{definition}

\begin{restatable}[SortedIx\_iff]{theorem}{sortedixiff}
  For all lists $l$,
  \begin{equation*}
    \var{Sorted}~l \iff \var{SortedIx}~l
  \end{equation*}
\end{restatable}
\begin{proof}
  See \cref{appendix:sorted}.
\end{proof}

However, a sorting function cannot produce just any sorted list; the
output must also contain the same elements as the original list. To
capture this, we could use the \var{Permutation} relation defined in
the Coq standard library. This relation holds on two lists when they
contain all the same elements, and is defined inductively:
\begin{definition}[Permutation]
  \label{def:permutation}
  \begin{gather*}
    \infer[\var{perm\_nil}]{\var{Permutation}~[]~[]}{}
    \\
    \allowbreak
    \\
    \infer[\var{perm\_skip}]{\var{Permutation}~(x :: l)~(x :: l')}{%
      \var{Permutation}~l~l'
    }
    \\
    \allowbreak
    \\
    \infer[\var{perm\_swap}]{\var{Permutation}~(x :: y :: l)~(y :: x :: l)}{}
    \\
    \allowbreak
    \\
    \infer[\var{perm\_trans}]{\var{Permutation}~l~l''}{
      \var{Permutation}~l~l' & \var{Permutation}~l'~l''
    }
  \end{gather*}
\end{definition}
While this definition is useful, proofs that do induction directly on
the evidence of \var{Permutation} can sometimes get stuck on the
transitive case, hence the remark in the introduction about the
usefulness of \cref{thm:stablesort_unique}.

Before we go further, we should clarify what we mean by the word
\textit{permutation}, as it has two related meanings that are often
used interchangeably. We sometimes say that two lists are permutations
of one other if they contain all of the same elements, perhaps in a
different order, like the lists $l =
[\var{zebra},\ \allowbreak\var{appliance},\ \allowbreak\var{kite}]$
and $l' =
[\var{appliance},\ \allowbreak\var{kite},\ \allowbreak\var{zebra}]$
However, in mathematics we sometimes say that a permutation is
actually the bijection on the indices that defines the correspondence
between the two lists. In this case, we could map indices in $l$ to
indices in $l'$ by mapping $0 \mapsto 2$, mapping $1 \mapsto 0$, and mapping $2 \mapsto
0$. $p$ is the permutation that relates $l$ to $l'$ if for all indices
$i$ and $j$,
\begin{equation*}
  l_i = l'_{p(i)}
\end{equation*}
For consistency with Coq's built-in \var{Permuation} relation, in a
situation like this we will say that $l$ and $l'$ are permutations of
one other, and that $p$ is the \textit{permutation function} that
relates $l$ to $l'$.

This suggests an alternative definition for \var{Permutation}:
\begin{definition}[PermutationEx]
  \label{def:permutation_ex}
  Two lists $l$ and $l'$ are permutations of one another when there
  exists a permutation function $p$ on $n = \var{length}~l$ elements
  such that $p(l) = l'$.
\end{definition}
Here we are overloading the notation $p(l)$ to denote the result of
shuffling a list~$l$ according to a permutation function~$p$. To
define this precisely, we have to first define how we will represent
permutation functions in Coq.

A permutation function on $n$ elements maps indices to indices, and so
has type $\nat_{<n} \to \nat_{<n}$. In Coq, we can represent a finite
function as a list of its outputs: $f : \nat_{<n} \to A$ is represented
by the list $[f(0),~f(1),\allowbreak\ldots,~f(n-1)]$. Applied to
permutations, this means that the permutation~$p$ mentioned above
would be represented by $[2,~1,~0]$. This is consistent with
``one-line'' notation for permutations, in which a permutation on $n$
elements is represented by its result when applied to the
list~$[0,~2,\ldots,n-1]$. Now we can define the action of a permutation on
a list.
\begin{equation*}
  p(l) \coloneq \var{map}~(\lambda i. l_i)~p
\end{equation*}
We implement this in Coq using an \var{apply} function. The definition
is slightly complicated by the fact that \var{nth} takes an extra
parameter to return in case $i$ is out of bounds.
\begin{lstlisting}
Definition apply p l : list A :=
  match l with
  | [] => []
  | d :: _ => map (fun i => nth i l d) p
  end.
\end{lstlisting}

However, not any list of natural numbers represents a valid
permutation function, because a permutation must be a bijection. We
can define the property $\var{PermFun}~n~p$, which holds when $p :
\var{list}~\nat$ represents a permutation function on $n$ elements.
\begin{definition}[PermFun]
  A list~$p$ of natural numbers is a permutation function on $n$
  elements if for all natural numbers $i$, $i \in p \iff i < n$, and if
  $p$ contains no duplicate elements.
\end{definition}

\begin{restatable}[PermutationEx\_iff]{theorem}{permutationexiff}
  \label{thm:permutationex_iff}
  For all lists $l$ and $l'$,
  \begin{equation*}
    \var{Permutation}~l~l' \iff \var{PermutationEx}~l~l'
  \end{equation*}
\end{restatable}
\begin{proof}
  See \cref{appendix:perm}.
\end{proof}

@@

\begin{restatable}[PermutationCount\_iff]{theorem}{permutationcountiff}
  \label{thm:permutationcount_iff}
  For all lists $l$ and $l'$,
  \begin{equation*}
    \var{Permutation}~l~l' \iff \var{PermutationCount}~l~l'
  \end{equation*}
\end{restatable}
\begin{proof}
  See \cref{appendix:perm}.
\end{proof}

With this we can define the specification for sorting functions:
\begin{definition}[Sort]
  \label{def:sort}
  A function $f : \var{list}~A \to \var{list}~A$ is a sorting function
  when for all inputs $l$,
  \begin{equation*}
    \var{Sorted}~(f~l) \land \var{Permutation}~(f~l)~l
  \end{equation*}
\end{definition}

@@

Now that we've defined sorts, we can consider stable sorts. Stable
sorts must preserve the relative orders of ``equivalent'' elements.
This equivalence comes from the ordering we are using: any total
decidable preorder also give us a decideable equivalence relation $\equiv$,
where $x \equiv y$ when $x \le y$ and $x \ge y$.

@@

To define stable permutations, we can tweak each of the definitions of
permutations mentioned above. For example, we can modify our
function-based definition.

@@

We can also modify the inductive definition of permutations to respect
equivalence. The problematic rule is \var{perm\_swap} because we could
potentially swap two equivalent elements. We will restrict it, so that
you can only swap $x$ and $y$ when $x \nequiv y$.

@@

Finally, we can say that if two lists are in a stable permutation, and
for some $x$ we examine only the the elements equivalent to $x$, then
those elements should appear in the same order in both lists.
\begin{definition}[StablePerm]
\item Two lists $l$ and $l'$ are stable permutations of one another if
  for every $x$,
  \begin{equation*}
    \var{filter}~(\lambda y. y \equiv x)~l = \var{filter}~(\lambda y. y \equiv x)~l'
  \end{equation*}
\end{definition}
In a sense, this is a stricter version of
\cref{def:permutation_count}, because that can be rephrased as
$\var{filter}~(\lambda y. y = x)~l = \var{filter}~(\lambda y. y = x)~l'$.

\begin{restatable}[StablePermEx\_iff]{theorem}{stablepermexiff}
  For all lists $l$ and $l'$
  \begin{equation*}
    \var{StablePerm}~l~l' \iff \var{StablePermEx}~l~l'
  \end{equation*}
\end{restatable}

\begin{restatable}[StablePermInd\_iff]{theorem}{stablepermindiff}
  For all lists $l$ and $l'$
  \begin{equation*}
    \var{StablePerm}~l~l' \iff \var{StablePermInd}~l~l'
  \end{equation*}
\end{restatable}

@@

This means that any two stable sorts must produce the same results for
every input, and are thus interchangeable.
\stablesortunique

This can often be extended even to unstable sorting algorithms.
Remember, it isn't always true that $x \equiv y$ implies $x = y$. For
example, if we are sorting strings by their first character, then
$\text{\texttt{fit}} \equiv \text{\texttt{fog}}$, but clearly
$\text{\texttt{fit}} \neq \text{\texttt{fog}}$. When we do have that the
equivalence ($\equiv$) implies Leibniz equality ($=$), we say that we have
an \textit{order}, as we have when we are sorting strings using all of
their characters. Under an order, as opposed to a preorder, equivalent
elements are actually indistinguishable, so you can't tell if they
have been swapped.

@@

Least-significant digit (LSD) radix sort is a simple radix sort that
is used to lexicographically sort strings of equal length. We can
define what it means to sort lexicographically:

\begin{definition}[lex\_le]
  Given a preorder~$\le$ on a type $A$, we define the preorder $\lexle$
  on $\var{list}~A$.
  \begin{gather*}
    \infer[\var{lex\_le\_nil}]{[] \lexle ys}{}
    \\
    \\
    \infer[\var{lex\_le\_cons\_lt}]{(x :: xs) \lexle (y :: ys)}{%
      x < y
    }
    \\
    \\
    \infer[\var{lex\_le\_cons\_eq}]{(x :: xs) \lexle (y :: ys)}{%
      x \equiv y & xs \lexle ys
    }
  \end{gather*}
\end{definition}

LSD radix sort operates by conceptually placing all the strings in a
matrix, and then stably sorting on each column, from the end of the
string to the front (\cref{fig:lsd_radixsort}). This works because if
two strings match on a column, the stable sort ``defers'' to the
preexisting ordering, and that preexisting ordering was determined by
sorting by the tail of the strings. This mirrors the way a
lexicographic ordering ``defers'' to the rest of a string when two
strings match on earlier characters.

\begin{figure}
  \centering
  \begin{tt}
    \setlength{\tabcolsep}{0pt}
    \begin{tabular}{c}
    fit \\
    rib \\
    fly \\
    vat \\
    age \\
    mix \\
    row \\
    fog
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    ri&b& \\
    ag&e& \\
    fo&g& \\
    fi&t& \\
    va&t& \\
    mi&x& \\
    fl&y& \\
    ro&w&
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    v&a&t \\
    a&g&e \\
    r&i&b \\
    f&i&t \\
    m&i&x \\
    f&l&y \\
    f&o&g \\
    r&o&w
    \end{tabular}
    $\rightarrow$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &a&ge \\
    &f&it \\
    &f&ly \\
    &f&og \\
    &m&ix \\
    &r&ib \\
    &r&ow \\
    &v&at
    \end{tabular}
  \end{tt}
  \caption{LSD radix sort in-place}
  \label{fig:lsd_radixsort}
\end{figure}

In a functional setting, we can define an LSD radix sort elegantly by
repeatedly moving the last column to the front of the matrix and
sorting on the first column (\cref{fig:lsd_radixsort_func}). First, we
define \var{rrot}, which moves the last element of a list to the
front, and relies on \var{init}, which drops the last element of a
list. Note that in order to ensure that $\var{last}$ is total, it
takes an extra argument to return in the case of an empty list.
\begin{lstlisting}
Definition rrot (l: list A) : list A :=
  match l with
  | [] => []
  | hd :: tl => last l hd :: init l
  end.

Fixpoint init l : list A :=
  match l with
  | [] | [_] => []
  | hd :: tl => hd :: init tl
  end.
\end{lstlisting}

Now we can define \var{radixsort}, taking the number of columns $n$ as
a parameter.
\begin{equation*}
  \var{radixsort}~m~n \coloneq (\var{hdsort} \comp \var{map}~\var{rrot})^n~m
\end{equation*}
We use the notation $f^i z$ for $\var{rep}~f~i~z$, which implements
repeated function application.

\begin{figure}
  \centering
  \begin{tt}
    \setlength{\tabcolsep}{0pt}
    \begin{tabular}{c}
    fit \\
    rib \\
    fly \\
    vat \\
    age \\
    mix \\
    row \\
    fog
    \end{tabular}
    $\overset{\var{rot}}{\rightarrow}$
    \begin{tabular}{c}
    tfi \\
    bri \\
    yfl \\
    tva \\
    eag \\
    xmi \\
    wro \\
    gfo
    \end{tabular}
    $\overset{\var{sort}}{\rightarrow}$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &b&ri \\
    &e&ag \\
    &g&fo \\
    &t&fi \\
    &t&va \\
    &x&mi \\
    &y&fl \\
    &w&ro
    \end{tabular}
    $\overset{\var{rot}}{\rightarrow}$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    ibr \\
    gea \\
    ogf \\
    itf \\
    atv \\
    ixm \\
    lyf \\
    owr
    \end{tabular}
    $\overset{\var{sort}}{\rightarrow}$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &a&tv \\
    &g&ea \\
    &i&br \\
    &i&tf \\
    &i&xm \\
    &l&yf \\
    &o&gf \\
    &o&wr
    \end{tabular}
    $\overset{\var{rot}}{\rightarrow}$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    vat \\
    age \\
    rib \\
    fit \\
    mix \\
    fly \\
    fog \\
    row
    \end{tabular}
    $\overset{\var{sort}}{\rightarrow}$
    \begin{tabular}{c>{\columncolor[gray]{0.9}}cc}
    &a&ge \\
    &f&it \\
    &f&ly \\
    &f&og \\
    &m&ix \\
    &r&ib \\
    &r&ow \\
    &v&at
    \end{tabular}
  \end{tt}
  \caption{LSD radix sort using rotations}
  \label{fig:lsd_radixsort_func}
\end{figure}

Here \var{hdsort} is an insertion sort
(\cref{appendix:insertion_sort}) that only examines the first
character of each string. We can implement this by defining orderings
that compare elements using some key function:

@@

For both properties, our approach will be to prove some invariant that
holds after $j$ iterations. For \var{StablePerm}, we can say that
after $j$ iterations of rotating and sorting, the matrix should be a
stable permutation of the original matrix had we only rotated $j$
times without sorting.
